Identification of rice genes which respond to saline stress from co-expression network analysis

### 2020-10-05 camila/Report/article_CR.tex

\subsection{Modules association to phenotypic traits}
The phenotypic traits under study are shoot $K^+$ content, root biomass and shoot biomass. Figure \ref{fig:pdata} suggests that there are significant differences in the values of these phenotypic traits between stress and control conditions. This supports the working hypothesis that these three variables represent tolerance-associated traits in rice under salt stress.

@@

The proposed methodology follows the principles of WGCNA. Starts from the level of thousands of genes, identifies biologically interesting gene modules, and identifies key genes in the treatment pathways for further validation. In WGCNA the whole workflow is structured in a modular way. The proposed methodology takes advantage of this property modifying and exploring new techniques inside some processes.\\

An important modification to the WGCNA methodology is in the pre-processing of the input data. Our approach is designed to integrate expression data measured under two different conditions (control and treatment), instead of working with them independently or considering a single condition. For this purpose, an approach similar to that proposed in~\cite{du2019network} is used, where the control and treatment data are compiled in a single matrix using the log fold change measure. Thus, the input to construct the co-expression network is not the expression data but the changes in the expression levels from one condition to the other, allowing to capture the signal of changes caused by the treatment.\\

The most innovative feature of the proposed methodology is the module detection technique. 
The co-expression network is calculated as in WGCNA until the scale-free network is obtained. In our approach, this network is used to apply the HLC algorithm, which is capable of detecting overlapping communities. The usual approximation and its different variations, only focuses on disjoint (non-overlapping) communities, but it is well known that communities in real networks are overlapping. Thus the approach presented in this work can be seen as a generalization of the previous approaches.\\

>>>>

\subsection{Modules association to phenotypic traits}
The phenotypic traits under study are shoot $K^+$ content, root biomass and shoot biomass. Figure \ref{fig:pdata} suggests that there are significant differences in the values of these phenotypic traits between stress and control conditions. This supports the working hypothesis that these three variables represent tolerance-associated traits in rice under salt stress.\\

@@

The proposed methodology is inspired by the workflow suggested by the WGCNA methodology~\cite{langfelder2008wgcna}. The main steps are the preprocessing of the gene expression data, the construction of a co-expression network, the detection of modules within the network, relate the modules with external information (e.g. phenotypic data) and an enrichment of the identified key genes. The methodology is structured in a modular way, which allows modifying and exploring different techniques in each of the steps of the process. \\

%The proposed methodology follows the principles of WGCNA. Starts from the level of thousands of genes, identifies biologically interesting gene modules, and identifies key genes in the treatment pathways for further validation. In WGCNA the whole workflow is structured in a modular way. The proposed methodology takes advantage of this property modifying and exploring new techniques inside some processes.\\

The proposed approach is designed to integrate expression data measured under two different conditions (control and treatment), unlike the usually co-expression-based approaches which work with both conditions independently or consider a single condition. For this purpose, an approach similar to that proposed in~\cite{du2019network} is used, where the control and treatment data are compiled in a single matrix using the log fold change measure. Thus, the input to construct the co-expression network is not the expression data but the changes in the expression levels from one condition to the other, allowing to capture the signal of changes caused by the treatment.\\

Another important innovation in the proposed methodology is the module detection technique. The co-expression network is calculated as in WGCNA until the scale-free network is obtained. This network is then used to apply the HLC algorithm, a clustering technique capable of detecting overlapping communities. Several approaches of module detection from gene expression have been proposed and were evaluated in~\cite{saelens2018comprehensive}. Most of them focuses only on disjoint (non-overlapping) communities and the described techniques dealing with overlap are not clustering but biclustering and decomposition methods. It is well known that communities in real networks are overlapping. Thus the approach presented in this work can be seen as a generalization of the previous approaches.\\
